---
title: "TarMetCL"
author: "Hongchao Ji"
date: "2021/3/12"
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Targeted metabolic analyses based on mass spectrometry}
  %\VignetteKeywords{mass spectrometry, metabolomics}
  %\VignetteEncoding{UTF-8}
  %\VignettePackage{TarMet}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This is a guide for using TarMet with R scripts.

# Targeted metabolite analysis
## Preparation
```{r preparation, message = FALSE, echo = TRUE, warning=FALSE}
library(TarMet)

data("isotopes", package = "enviPat")
data("adducts", package = "enviPat")

# path of raw data
rawfiles <- list.files(system.file('rawfiles', package = 'TarMet'), full.names = TRUE)  # path of data files
samples <- sapply(strsplit(rawfiles, '/'), function(s) s[length(s)])                    # name of data files
rawDataset <- lapply(rawfiles, LoadData)                                                # load raw data

# path of config file
config <- read.csv(system.file('data/targets.csv', package = 'TarMet'))

# global parameters
fineness <- 'Medium'  # fineness of peak detection of MassSpecWavelet
resolution <- 50000   # resolution of your MS

shift <- 20           # parameter for alignment
segment <- 20         # parameter for alignment

# select the referance sample
sampleInd <- 1        # choose a reference file
```

## Set targeted metabolite
```{r targeted metabolite, message = FALSE, echo = TRUE, warning=FALSE}
i = 1                 # use the first metabolite

# set the parameters for the specific metabolite
threshold <- 0.01
name <- config$name[i]
formula <- config$formula[i]
mz <- config$mz[i]
  
# adduct type
if (is.na(config$adduct[i])){
  adduct <- 'M+H'
} else {
  adduct <- config$adduct[i]
}
  
  # ppm for EIC generation
if (is.na(config$ppm[i])){
  ppm <- 100
} else {
  ppm <- config$ppm[i]
}
  
  # scales for peak detection
if (is.na(config$scale[i])){
  scale <- 5
} else {
  scale <- config$scale[i]
}
  
  # start of retention time for EIC generation 
if (is.na(config$rtmin[i])){
  rtmin <- 0
} else {
  rtmin <- config$rtmin[i]
}

# end of retention time for EIC generation 
if (is.na(config$rtmax[i])){
  rtmax <- max(rawDataset[[1]]$times)
} else {
  rtmax <- config$rtmax[i]
} 
  
# minimum peak height of a peak
if (is.na(config$height[i])){
  height <- 0
} else {
  height <- config$height[i]
}
  
# minimum SNR of a peak
if (is.na(config$snr[i])){
  snr <- 5
} else {
  snr <- config$snr[i]
}
```

## Calculate targeted m/z value
```{r mz calculation, message = FALSE, echo = TRUE, warning=FALSE}
pattern <- getIsotopicPattern(formula, adduct, threshold, resolution)
mzs <- pattern[,1]
targetMzRanges <- getMzRanges(mzs, resolution=resolution, ppm=ppm)
```

## Targeted EIC generation
```{r eic generation, message = FALSE, echo = TRUE, warning=FALSE}
targetEICs <- list()
rtranges <- c(rtmin, rtmax)
for (j in seq_along(rawfiles)){
  targetEICs[[j]] <- getMzEICs(rawDataset[[j]], rtranges=rtranges, mzranges=targetMzRanges, baseline=FALSE, smooth=FALSE)
}
  
theoretical <- as.numeric(pattern[,2])
targetPeaks <- getIsotopicPeaks(targetEICs[[sampleInd]], SNR.Th=snr, peakScaleRange=scale, peakThr=height, theoretical=theoretical, fineness=fineness)
plotEICs(targetEICs[[sampleInd]], targetPeaks)
```

## Display peak detection results
```{r peak info, message = FALSE, echo = TRUE, warning=FALSE}
outputPeakInfo <- targetPeaks$PeakInfo
knitr::kable(outputPeakInfo, format = "markdown")
```

## Alignment
```{r alignment, message = FALSE, echo = TRUE, warning=FALSE}
alignedEICs <- getAlignedEICs(targetEICs, sampleInd, shift, segment, align=TRUE)
WhtoPlot <- 1  # plot the monoisotopic feature of the targeted metabolite of all samples
EicstoPlot <- lapply(alignedEICs, function(eics){
  eics[[WhtoPlot]]
})
names(EicstoPlot) <- samples
plotEICs(EicstoPlot, rtrange=c(outputPeakInfo$Start[WhtoPlot], outputPeakInfo$End[WhtoPlot]))
```

## Display peak area of all samples
```{r alignment results, message = FALSE, echo = TRUE, warning=FALSE}
whPeak <- which.max(outputPeakInfo$Similarity)  # Using the feature with most similar isotopic pattern
outputSamplesArea <- getQuantifiedResult(alignedEICs, outputPeakInfo$Start[whPeak], outputPeakInfo$End[whPeak])
colnames(outputSamplesArea) <- samples
knitr::kable(outputSamplesArea, format = "markdown")
```

## Display barplot of isotopes
```{r isotope barplot, message = FALSE, echo = TRUE, warning=FALSE}
plotStackBar(outputSamplesArea)
```

# Isotope tracer analysis
## Preparation
```{r isotope tracer data, message = FALSE, echo = TRUE, warning=FALSE}

if (!'AssayRdata' %in% installed.packages()){
  install.packages(
    "https://gitlab.com/jimiwills/assay.R/raw/master/AssayR_0.1.5.tar.gz", 
    repos = NULL, type = "source")
  install.packages(
    "https://gitlab.com/jimiwills/assay.R/raw/2190202fd4bc0325421bb10c4ec42089d45e1952/AssayRdata_0.1.3.tar.gz", 
    repos = NULL, type = "source")
}

rawfiles <- list.files(system.file("extdata", package = "AssayRdata"), full.names = TRUE)
samples <- sapply(strsplit(rawfiles, '/'), function(s) s[length(s)])
rawDataset <- lapply(rawfiles, LoadData)
```

## Set targeted metabolite
```{r tracer target, message = FALSE, echo = TRUE, warning=FALSE}
name <- 'glucose'
formula <- 'C6H12O6'
adduct <- 'M-H'

ppm <- 50
scale <- 10
rtmin <- 0
rtmax <- max(rawDataset[[1]]$times)
height <- 0
snr <- 10
```

## Calculate m/z value
```{r tracer m/z, message = FALSE, echo = TRUE, warning=FALSE}
tracer_element <- 'C'       # type of element
tracer_isotope <- '13C'     # type of isotope tracer
tracer_number <- 6          # maximum number of isotope tagger
mzs <- getMzWithTracer(formula, adduct, tracer_element, tracer_isotope, tracer_number)
targetMzRanges <- getMzRanges(mzs, resolution=resolution, ppm=ppm)
```

## Targeted EIC generation
```{r tracer eic generation, message = FALSE, echo = TRUE, warning=FALSE}
targetEICs <- list()
rtranges <- c(rtmin, rtmax)
for (j in seq_along(rawfiles)){
  targetEICs[[j]] <- getMzEICs(rawDataset[[j]], rtranges=rtranges, mzranges=targetMzRanges, baseline=FALSE, smooth=FALSE)
}
  
theoretical <- as.numeric(pattern[,2])
targetPeaks <- getIsotopicPeaks(targetEICs[[sampleInd]], SNR.Th=snr, peakScaleRange=scale, peakThr=height, theoretical=theoretical, fineness=fineness)
plotEICs(targetEICs[[sampleInd]], targetPeaks)
```

## Display peak information
```{r tracer peakinfo, message = FALSE, echo = TRUE, warning=FALSE}
outputPeakInfo <- targetPeaks$PeakInfo
knitr::kable(outputPeakInfo, format = "markdown")
```

## Set peak bounds manually
Set peak bounds manually to combine peak 1 and peak 2
```{r user info, message = FALSE, echo = TRUE, warning=FALSE}
targetRtLeft <- 900.492
targetRtRight <- 976.229
targetRtAxis <- 928.768
userInfo <- data.frame(Name='User', Position=targetRtAxis, Start=targetRtLeft, End=targetRtRight)
outputPeakInfo <- rbind(targetPeaks$PeakInfo[,1:ncol(userInfo)], userInfo)
knitr::kable(outputPeakInfo, format = "markdown")
```

## Calculate peak area
```{r peakinfo, message = FALSE, echo = TRUE, warning=FALSE}
userArea <- round(getArea(targetEICs[[sampleInd]], targetRtLeft, targetRtRight), 2)
outputPeakArea <- {
    res <- targetPeaks$PeakArea
    res <- cbind(rownames(res), res)
    res <- cbind(res, userArea, round(userArea/sum(userArea),3))
    colnames(res)[1] <- 'mzRange'
    colnames(res)[(ncol(res)-1) : ncol(res)] <- c('User','Abundance')
    res
}

knitr::kable(outputPeakArea, format = "markdown")
```

## Alignment
```{r tracer alignment, message = FALSE, echo = TRUE, warning=FALSE}
alignedEICs <- getAlignedEICs(targetEICs, sampleInd, shift, segment, align=TRUE)
WhtoPlot <- 7   # plot the M+6 feature of the targeted metabolite of all samples
EicstoPlot <- lapply(alignedEICs, function(eics){
  eics[[WhtoPlot]]
})
names(EicstoPlot) <- samples
plotEICs(EicstoPlot, rtrange=c(outputPeakInfo$Start[WhtoPlot], outputPeakInfo$End[WhtoPlot]))
```

## Display peak area of all samples
```{r tracer alignment results, message = FALSE, echo = TRUE, warning=FALSE}
whPeak <- which(outputPeakInfo$Name == 'User')
outputSamplesArea <- getQuantifiedResult(alignedEICs, outputPeakInfo$Start[whPeak], outputPeakInfo$End[whPeak])
colnames(outputSamplesArea) <- samples
knitr::kable(outputSamplesArea, format = "markdown")
```

## Display barplot of isotopes
```{r tracer isotope barplot, message = FALSE, echo = TRUE, warning=FALSE}
plotStackBar(outputSamplesArea)
```

